A NetdiffuseR Tutorial: Estimating Network Influences on Behavior within the Diffusion Paradigm

### 2024-12-05 example.qmd
Working on the draft

---
title: "A NetdiffuseR Tutorial: Estimating Network Influences on Behavior within the Diffusion Paradigm"
format:
  pdf:
    classoption:
      - twocolumn
keep-tex: true
bibliography: bibliography.bib
author: 
  - name: George G. Vega Yon
    affiliation: University of Utah
  - name: Aníbal L. Olivera M.
    affiliation: Universidad del Desarrollo
  - name: Thomas W. Valente
    affiliation: University of Southern California
# include-in-header: header.tex
---

```{r}
#| label: setup
#| echo: false
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(width = 45)
```

@@

There are other measures needed to estimate diffusion network effects. Network thresholds (Valente, 1996) are the proportion or number of adopters in one’s network when an individual adopts. Some people innovate with none or few adopters in their network adopting whereas others wait until a majority or all adopt. Network thresholds provide a means to estimate the classic two-step flow hypotheses (Katz & Lazarsfeld, 1957) and have been shown to moderate health promotion impacts (Valente & Saba, 1998). The threshold distribution drives the diffusion rate and prevalence (Valente & Vega Yon, 2020). Other diffusion network measures are Moran’s I, a measure of contagion, network susceptibility, network influence (Valente, et al, 2015), the hazard rate, and Bass model parameter coefficients (Bass, 1969; Valente, 1993).
# The netdiffuseR package

@@

In a nutshell, netdiffuseR provides methods for simulating, processing, analyzing, and visualizing diffusion networks. Among its key functions are the survey_to_diffnet function rdiffnet function, and the exposure function. The survey_to_diffnet function provides an interface to read in tabular data (in R data.frame format) corresponding to survey data containing network nominations and time of adoption [TOA]. The function makes it easy to process survey data, pass multiple checks, and build the adoption and cumulative adoption matrices. The rdiffnet function allows simulating diffusion networks using the threshold model and, since version 1.3.0, simulating multi-diffusion and disadoption. Users can simulate diffusion processes on pre-defined networks or random graphs in a computationally efficient way. Finally, the exposure function–the workhorse of the R package–provides a flexible framework for calculating various types of exposure statistics, including A, B, C. In the following section of the paper, we will provide a walk thorough to these functions.
# A tutorial on netdiffuseR
## Analyzing existing data (classical models)

@@

[PROTIP] You can always inspect what datasets are included in an R package using the `data()` function. For example, in netdiffuseR, you can use `data(package="netdiffuseR")`.

@@

[PROTIP] The Moran’s I statistic is computed with the inverse of the geodesic matrix (shortest path matrix) but can be computed using other statistics using the `moran` function implemented in netdiffuseR.

Of the observed outcomes, it is noteworthy that (a) the diffusion process is apparent in the figure, with adoptions occurring from core to periphery of the graph and (b) Moran’s I shows a high-level of spatial autocorrelation between periods four and ten. Other visualization functions include `plot_diffnet2`, `plot_threshold`, `plot_adopters`, and `plot_infectsuscept`. We invite the reader to learn more about them in the manual of the netidffuseR R package.
## Computing exposures

@@

$$
An equation with the math for the homophilic exposure.
$$

The following code block illustrates how we can use `vertex_covariate_compare` to generate an alternative graph for computing age-homophilic exposure:

@@

In sum, the `exposure` function in the package provides a great deal of flexibility for calculating a wide range of exposures. Computing exposures is a fundamental task in the study of diffusion networks. A more in-depth treatment of network exposures is given in [CITE TOM PAPER’S ABOUT EXPOSURE].
## Simulating diffusion networks

@@

As a first example, you can give as input only the number of nodes and the time steps. The function then constructs as default a scale-free graph with random rewiring for each time step, sets the proportion of initial adopters to 0.05, and imputes a random number between 0 and 1 as the fractional threshold value for each node. This example shows the default settings:

```{r}
#| label: simulating-adoption-process-1

n          <- 200				# Number of nodes
t           <- 10				# Time steps

diffnet_1 <- rdiffnet(
  n,
  t,
  seed.nodes = "random",
  seed.p.adopt = 0.05,
  seed.graph = "scale-free",
  rgraph.args = list(),
  rewire = TRUE,
  rewire.args = list(),
  threshold.dist = runif(n),
  exposure.args = list(),
  name = "A diffusion network",
  behavior = "Random contagion",
  stop.no.diff = TRUE,
  disadopt = NULL
)

diffnet_1
```
As stated before, for the simulation of a single behavior you can specify the threshold distribution —giving `threshold.dist` as a function, a numeric scalar, or a vector of $n$ elements—, the proportion of early adopters —`seed.p.adopt` as a numeric scalar—, the nature of those nodes based on degree centrality or specifying the actual nodes that will acts as seeds —`seed.nodes` as a character scalar o a vector, respectively—, and the name of the behavior —`behavior` as a character scalar—. For example:
```{r}
#| label: simulating-adoption-process-2

diffnet_2 <- rdiffnet(n, t, seed.p.adopt = 0.1, threshold.dist = runif(n, .3,.5), seed.nodes = 'central', behavior = 'tobacco')

diffnet_2
```

Alternatively, you can give an already constructed graph as an input:

```{r}
#| label: simulating-adoption-process-3

set.seed(121)

graph      <- rgraph_ws(n, 10, p=.3)   # Creating a graph based on watts-strogatz model

# Run the simulation
diffnet_3 <- rdiffnet(seed.graph = graph, t = t , seed.p.adopt = 0.1, threshold.dist = runif(n, .3,.5))
diffnet_3

```

@@

netdiffuseR also can handle adoption processes for more than one innovation/behavior. To select a multiadoption process, you only must give a `list` as the `seed.p.adopt` parameter. Therefore the simplest way to run `rdiffnet` with multiple behaviors could be:
```{r}
#| label: simulation-multiadoption-process-4
diffnet_4 <- rdiffnet( n,  t, seed.p.adopt = list(0.1, 0.15))
diffnet_4
```
Now, the characteristics of the behaviors being simulated can be given together, all sharing the same specifications,
```{r}
#| label: simulation-multiadoption-process-5
diffnet_5 <- rdiffnet( n,  t, seed.p.adopt = list(0.1, 0.15), threshold.dist = runif(n, .3,.5),
                       seed.nodes = 'central', behavior = 'tobacco')
diffnet_5
```
and in that case, for two behaviors, they are named `‘behavior_name’`_1 and `‘behavior_name’`_2; or you can specify different characteristics for each behavior being simulated:
```{r}
#| label: simulation-multiadoption-process-6
diffnet_6 <- rdiffnet(
 n, t, seed.p.adopt = list(0.1, 0.15),
 threshold.dist = list(runif(n, .3,.5), runif(n, .2,.4)),
 seed.nodes = list('central', 'random'),
 behavior = list('tobacco', 'alcohol')
 )

diffnet_6
```
In this particular example, we ran `rdiffnet` with most of the parameters being `list`. The reader is invited to look at other types of possible inputs running `?rdiffnet`.

Same as before, you can give a specific network as input.
```{r}
#| label: simulation-multiadoption-process-7

# Run the simulation with the same parameters as before but specifying different initial 
# proportion of adopters for each behavior
diffnet <- rdiffnet(seed.graph = graph, t = t , seed.p.adopt = list(0.1, 0.15),
                    threshold.dist = runif(n, .3,.5))
diffnet

```
## Disadoption

@@

graph <- rgraph_ws(n, t, p=.3)

@@

diffnet_7  <- rdiffnet(seed.graph = graph, t = t, disadopt = random_dis,
                               seed.p.adopt = 0.1)
diffnet_7 
```
And, as a second example, we can build a disadoption function that prevents the nodes from adopting more than one behavior if `rdiffnet` simulates two of them:

@@

diffnet_8 <- rdiffnet(n, t, disadopt = one_only, seed.p.adopt = list(0.1, 0.1))
diffnet_8
```
# Conclusion

@@

>>>>

---
title: "A NetdiffuseR Tutorial: Estimating Network Influences on Behavior within the Diffusion Paradigm"
format:
  acs-pdf:
    keep-tex: true  
keep-tex: true
bibliography: bibliography.bib
author: 
  - name: George G. Vega Yon
    affiliation:
      - name: University of Utah
    email: george.vegayon@utah.edu
  - name: Aníbal L. Olivera M.
    affiliation:
      name: Universidad del Desarrollo
  - name: Thomas W. Valente
    affiliation:
      name: University of Southern California
# include-in-header: header.tex
---

```{r}
#| label: setup
#| echo: false
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(digits=4)
```

@@

There are other measures needed to estimate diffusion network effects. Network thresholds (Valente, 1996) are the proportion or number of adopters in one’s network when an individual adopts. Some people innovate with none or few adopters in their network adopting whereas others wait until a majority or all adopt. Network thresholds provide a means to estimate the classic two-step flow hypotheses (Katz & Lazarsfeld, 1957) and have been shown to moderate health promotion impacts (Valente & Saba, 1998). The threshold distribution drives the diffusion rate and prevalence (Valente & Vega Yon, 2020). Other diffusion network measures are Moran’s I, a measure of contagion, network susceptibility, network influence (Valente, et al, 2015), the hazard rate, and Bass model parameter coefficients (Bass, 1969; Valente, 1993).

# The netdiffuseR package

@@

In a nutshell, netdiffuseR provides methods for simulating, processing, analyzing, and visualizing diffusion networks. Among its key functions are the survey_to_diffnet function rdiffnet function, and the exposure function. The survey_to_diffnet function provides an interface to read in tabular data (in R data.frame format) corresponding to survey data containing network nominations and time of adoption [TOA]. The function makes it easy to process survey data, pass multiple checks, and build the adoption and cumulative adoption matrices. The rdiffnet function allows simulating diffusion networks using the threshold model and, since version 1.3.0, simulating multi-diffusion and disadoption. Users can simulate diffusion processes on pre-defined networks or random graphs in a computationally efficient way. Finally, the exposure function–the workhorse of the R package–provides a flexible framework for calculating various types of exposure statistics, including A, B, C. In the following section of the paper, we will provide a walk thorough to these functions.

# A tutorial on netdiffuseR

## Analyzing existing data

@@

::: {.callout-tip}
You can always inspect what datasets are included in an R package using the `data()` function. For example, in netdiffuseR, you can use `data(package="netdiffuseR")`.
:::

@@

::: {.callout-tip}
The Moran’s I statistic is computed with the inverse of the geodesic matrix (shortest path matrix) but can be computed using other statistics using the `moran` function implemented in netdiffuseR.
:::

Of the observed outcomes, it is noteworthy that (a) the diffusion process is apparent in the figure, with adoptions occurring from core to periphery of the graph and (b) Moran’s I shows a high-level of spatial autocorrelation between periods four and ten. Other visualization functions include `plot_diffnet2`, `plot_threshold`, `plot_adopters`, and `plot_infectsuscept`. We invite the reader to learn more about them in the manual of the netidffuseR R package.

## Computing exposures

@@

$$
E_{it} = \frac{\sum_{i\neq j} W_{ij}a_j\|x_i - x_j\|^{-1}}{\sum_{i\neq j} W_{ij}\|x_i - x_j\|^{-1}}
$$

Where $\|x_i-x_j\|$ is L-2 norm between $i$ and $j$'s age. The following code block illustrates how we can use `vertex_covariate_compare` to generate an alternative graph for computing age-homophilic exposure:

@@

In sum, the `exposure` function in the package provides a great deal of flexibility for calculating a wide range of exposures. Computing exposures is a fundamental task in the study of diffusion networks. A more in-depth treatment of network exposures is given in [CITE TOM PAPER’S ABOUT EXPOSURE].

## Simulating diffusion networks

@@

The function has only two mandatory arguments: the number of nodes (`n`) and the time steps (`t`). By default, the function generates a scale-free graph with random rewiring for each time step, sets the proportion of initial adopters to 0.05, and sets a random uniform threshold value for each node. This example shows the default settings simulating a diffusion process with 200 nodes and 10 time steps:

```{r}
#| label: simulating-adoption-process-1
#| eval: false
diffnet_1 <- rdiffnet(
  # Mandatory arguments
  n = 200,
  t = 10,
  # Optional arguments
  seed.nodes     = "random",
  seed.p.adopt   = 0.05,
  seed.graph     = "scale-free",
  rgraph.args    = list(),
  rewire         = TRUE,
  rewire.args    = list(),
  threshold.dist = runif(n),
  exposure.args  = list(),
  name           = "A diffusion network",
  behavior       = "Random contagion",
  stop.no.diff   = TRUE,
  disadopt       = NULL
)
```

, which is equivalent to calling `rdiffnet(200, 10)`:

```{r}
#| label: simulating-adoption-process-call
#| eval: false
set.seed(331)
rdiffnet(200, 10)
```

Since the simulation process is inherently random, we use the function `set.seed` to ensure reproducibility.

::: {.callout-tip}
Although the threshold model is a deterministy model, since `rdiffnet` calls the function `runif` to generate the threshold values, the simulation process is stochastic. If we were using a fixed threshold value, the simulation would be deterministic.
:::

Currently, `rdiffnet` supports thresholds specified as a fixed value for all nodes, a vector with different values for each nodes, or a function to be called for each node. The proportion of early adopters (`seed.p.adopt`) is used in combination with the seed nodes (`seed.nodes`) parameters. Current supported values are `"random"`, `"central"`, or `"marginal"`; or, if passed as a vector, `rdiffnet` ignores `seed.p.adopt` and sets the initial adopters to be those nodes included in the `seed.nodes` vector.

The `rdiffnet` function also supports the use of a pre-defined graph as input. The following code shows how to simulate a diffusion process using a Watts-Strogatz graph:

```{r}
#| label: simulating-adoption-process-3
# Creating a graph based on watts-strogatz model
set.seed(121)
graph <- rgraph_ws(200, 10, p=.3)

# Run the simulation
diffnet_3 <- rdiffnet(
  t              = 10,
  seed.graph     = graph,
  seed.p.adopt   = 0.1,
  threshold.dist = runif(200, .3,.5)
  )

# Visualizing the diffusion process
plot_diffnet(diffnet_3)
```

@@

netdiffuseR also can handle adoption processes for more than one innovation/behavior. To select a multiadoption process, you only must give a `list` as the `seed.p.adopt` parameter. Therefore the simplest way to run `rdiffnet` with multiple behaviors could be:

```{r}
#| label: simulation-multiadoption-process-4
diffnet_4 <- rdiffnet(
  200, 10,
  seed.p.adopt = list(0.1, 0.15)
  )

diffnet_4
```

Now, the characteristics of the behaviors being simulated can be given together, all sharing the same specifications,

```{r}
#| label: simulation-multiadoption-process-5
diffnet_5 <- rdiffnet(
  200, 10,
  seed.p.adopt   = list(0.1, 0.15),
  threshold.dist = runif(200, .3,.5),
  seed.nodes     = 'central', 
  behavior       = 'tobacco'
  )

diffnet_5
```

and in that case, for two behaviors, they are named `‘behavior_name’`_1 and `‘behavior_name’`_2; or you can specify different characteristics for each behavior being simulated:

```{r}
#| label: simulation-multiadoption-process-6
diffnet_6 <- rdiffnet(
  200, 10,
  seed.p.adopt   = list(0.1, 0.15),
  threshold.dist = list(
    runif(200, .3,.5),
    runif(200, .2,.4)
  ),
  seed.nodes     = list('central', 'random'),
  behavior       = list('tobacco', 'alcohol')
 )

diffnet_6
```

In this particular example, we ran `rdiffnet` with most of the parameters being `list`. The reader is invited to look at other types of possible inputs running `?rdiffnet`.

Same as before, you can give a specific network as input. Besides passing fixed networks as we did with the small-world example one, the `rdiffnet` function also supports passing diffnet objects as input. When doing so, the function will use the graph of the diffnet object as the seed graph and will take the time argument as the number of timepoints included in the graph:

```{r}
#| label: simulation-multiadoption-process-7
# Run the simulation with the same parameters as before
# but specifying different initial proportion of adopters
# for each behavior
diffnet <- rdiffnet(
  seed.graph     = kfam_diffnet,
  seed.p.adopt   = list(0.1, 0.15),
  threshold.dist = runif(nvertices(kfam_diffnet), .3,.5)
  )

diffnet
```

To visualize the diffusion process when there's more than one behavior, we can use the `split_behaviors` function to split the diffnet object into a list of diffnet objects, one for each behavior. Then, we can use the `plot_adopters` function to visualize the diffusion process for each behavior:

```{r}
#| label: simulation-multiadoption-process-7-plot
diffnets <- split_behaviors(diffnet)
plot_adopters(diffnets[[1]])
```

## Disadoption

@@

graph <- rgraph_ws(200, 10, p=.3)

@@

diffnet_7  <- rdiffnet(
  seed.graph   = graph,
  t            = 10,
  disadopt     = random_dis,
  seed.p.adopt = 0.1
  )

diffnet_7 
```

And, as a second example, we can build a disadoption function that prevents the nodes from adopting more than one behavior if `rdiffnet` simulates two of them:

@@

diffnet_8 <- rdiffnet(
  200, 10,
  disadopt = one_only,
  seed.p.adopt = list(0.1, 0.1)
  )

diffnet_8
```

# Conclusion

@@

# References
